---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
[![R-CMD-check](https://github.com/mrc-ide/demogsurv/workflows/R-CMD-check/badge.svg)](https://github.com/mrc-ide/demogsurv/actions)
[![Coverage status](https://codecov.io/gh/mrc-ide/demogsurv/branch/master/graph/badge.svg)](https://codecov.io/github/mrc-ide/demogsurv?branch=master)
  
# demogsurv

The goal of demogsurv is to:

* Calculate common demographic indicators from household survey data, including child mortality, adult mortality, and fertility.
* Stratify indicators according to arbitrary age groups, calendar periods, birth cohorts, time before survey, or other survey variables (region, residence, wealth category, education, etc.).
* Efficiently prepare event / person-year datasets from survey data for other model-based analysis.
* Provide standard error estimates for all indicators via Taylor linearization or jackknife.
* Calculate sample covariance for indicators estimated from a single complex survey (e.g. time-series of child mortality estimates).
* Default parameters are specified for analysis of [DHS recode datasets](https://dhsprogram.com/data/) without fuss, but funciton implementation is flexible for use with other survey data.

For analysis of DHS data, the package interacts well with [`rdhs`](https://docs.ropensci.org/rdhs/). See the [vignette](https://github.com/mrc-ide/demogsurv/blob/master/vignettes/rdhs-integration.pdf) for an example.

## Installation

You can install the development version from [GitHub](https://github.com/mrc-ide/demogsurv) with:

``` r
# install.packages("devtools")
devtools::install_github("mrc-ide/demogsurv")
```

The package will be released on [CRAN](https://CRAN.R-project.org) in due course.


## Example

Load the package and example datasets created from [DHS Model Datasets](https://dhsprogram.com/data/model-datasets.cfm).
```{r}
library(demogsurv)

data(zzbr) # Births recode (child mortality)
data(zzir) # Individuals recode (fertility, adult mortality)
```

### Child mortality 

By default, the function `calc_nqx` calculates U5MR by periods 0-4, 5-9, and 10-14 
years before the survey. Before calculating mortality rates, create a binary variable
indicator whether a death occurred and a variable giving the date of death, placed 
0.5 months in the month the death occurred.

```{r u5mr}
zzbr$death <- zzbr$b5 == "no"      # b5: child is alive ("yes" or "no")
zzbr$dod <- zzbr$b3 + zzbr$b7 + 0.5
u5mr <- calc_nqx(zzbr)
u5mr
```

Note that `calc_nqx()` does **not** reproduce child mortality estimates 
produced in DHS reports. `calc_nqx()` conducts a standard demographic rate 
calculation based on observed events and person years within each age group
and then converts the cumulative hazard to survival probabilities. The standard
DHS indicator uses a rule-based approach to allocate child deaths and person 
years across age groups and proceeds by calculating direct probabilities of death 
in each age group (see [Rutstein and Rojas 2006](http://dhsprogram.com/pubs/pdf/DHSG1/Guide_to_DHS_Statistics_29Oct2012_DHSG1.pdf)). 
A function `calc_dhs_u5mr()` will reproduce the DHS calculation, but is not 
yet fully implemented.

Use the argument `by=` to specify factor variables by which to stratify the rate
calculation.

```{r u5mr by}
calc_nqx(zzbr, by=~v102) # by urban/rural residence
calc_nqx(zzbr, by=~v190, tips=c(0, 10)) # by wealth quintile, 0-9 years before
calc_nqx(zzbr, by=~v101+v102, tips=c(0, 10)) # by region and residence
```

The sample covariance or correlation matrix of the estimates can be obtained via `vcov()`.

```{r sample covariance}
vcov(u5mr)  # sample covariance
cov2cor(vcov(u5mr))  # sample correlation
```

Standard error estimation can be done via Taylor linearisation, unstratified 
jackknife, or stratified jackknife. Results are very similar.

```{r varmethod}
calc_nqx(zzbr, varmethod = "lin") # default is linearization
calc_nqx(zzbr, varmethod = "jkn") # stratified jackknife (varmethod = "jkn")

## Compare unstratified standard error estimates for linearization and jackknife
calc_nqx(zzbr, strata=NULL, varmethod = "lin")  # unstratified design
calc_nqx(zzbr, strata=NULL, varmethod = "jk1")  # unstratififed jackknife
```

To calculate different child mortality indicators (neonatal, infant, etc.), specify
different age groups over which to aggregate.

```{r NN PNN infant}
calc_nqx(zzbr, agegr=c(0, 1)/12)  # neonatal
calc_nqx(zzbr, agegr=c(1, 3, 5, 12)/12) # postneonatal
calc_nqx(zzbr, agegr=c(0, 1, 3, 5, 12)/12) # infant (1q0)
calc_nqx(zzbr, agegr=c(12, 24, 36, 48, 60)/12) # child (4q1)
calc_nqx(zzbr, agegr=c(0, 1, 3, 5, 12, 24, 36, 48, 60)/12) # u5mr (5q0)
```

Calculate annual ~5~q~0~ by calendar year (rather than years preceding survey).

```{r annual 5q0}
calc_nqx(zzbr, period=2005:2015, tips=NULL)
```

### Adult mortality

The function `calc_nqx()` can also used to calculate adult mortality indicators 
such as ~35~q~15~. First, the convenience function `reshape_sib_data()` transforms 
respondent-level data to a dataset with one row for each sibling reported. Then 
define a binary variable for whether the sibling is alive or dead.

```{r reshape sib}
zzsib <- reshape_sib_data(zzir)
zzsib$death <- factor(zzsib$mm2, c("dead", "alive")) == "dead"
```

Calculate ~35~q~15~ for the seven year period before the survey.
```{r sib death, warning=FALSE}
calc_nqx(zzsib, agegr=seq(15, 50, 5), tips=c(0, 7), dob="mm4", dod="mm8")
```

Calculate ~35~q~15~ by sex, replicating Table MM2.2.
```{r 35q15 by sex, warning=FALSE}
zzsib$sex <- factor(zzsib$mm1, c("female", "male"))  # drop mm2 = 3: "missing"
calc_nqx(zzsib, by=~sex, agegr=seq(15, 50, 5), tips=c(0, 7), dob="mm4", dod="mm8")
```

This calculation exactly reproduces the ~35~q~15~ estiamtes produced for Table MM2
for DHS reports. Additional functionality will be added in future for producing 
ASMRs (Table MM1), MMR, and PM (Table MM3) will be added in future.

### Fertility

The functions `calc_asfr()` and `calc_tfr()` calculate age-specific fertility 
rates and total fertility rate, respectively. The default calculation is by 
five-year age groups for three years before the survey, exactly reproducing 
the estimates produced in DHS reports.

```{r fertility}
## Replicate DHS Table 5.1.
## Total ASFR and TFR in 3 years preceding survey
calc_asfr(zzir, tips=c(0, 3))
calc_tfr(zzir)

## ASFR and TFR by urban/rural residence
reshape2::dcast(calc_asfr(zzir, ~v025, tips=c(0, 3)), agegr ~ v025, value.var = "asfr")
calc_tfr(zzir, by=~v025)
calc_tfr(zzir, by=~v025, varmethod="jkn")
```

Replicate fertility estimates stratified by various sociodemographic characteristics.

```{r table52}
## Replicate DHS Table 5.2
calc_tfr(zzir, ~v102)  # residence
calc_tfr(zzir, ~v101)  # region
calc_tfr(zzir, ~v106)  # education
calc_tfr(zzir, ~v190)  # wealth quintile
calc_tfr(zzir)  # total
```

Generate estimates stratified by both calendar period and time preceding survey.
```{r caltips}
calc_tfr(zzir, period = c(2010, 2013, 2015), tips=0:5)
```

Calculate ASFR by birth cohort.
```{r asfrcoh}
asfr_coh <- calc_asfr(zzir, cohort=c(1980, 1985, 1990, 1995), tips=NULL)
reshape2::dcast(asfr_coh, agegr ~ cohort, value.var = "asfr")
```

## To Do

* Refactor for single `calc_rate()` function used by mortality and fertility calculations.
* Add indirect indicators
* Implement DHS child mortality calculation (Rutstein and Rojas 2006).
* Handle married women only datasets.
* Calculate multiple aggregations of nqx and sample correlation of these (e.g. NN, PN, 1q0, 4q1, 5q0).
* Document everything...
* Develop, test, and create examples for non-DHS household surveys, e.g. MICS, WFS.
* Extend interface to allow specification of formula, variable name, or vectors.


* Add some basic data checking before rate calculation, good warnings, potentially
  automatic handling.
  [ ] Variable names all exist.
  [ ] No missing data in date variables.
  [ ] Episode start date is before episode end date.
  [ ] All birth history ids appear in respondent dataset.
