---
title: "Demographic indicator calcultions using `demogsurv`"
author: "Jeff Eaton and Bruno Masquelier"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Demographic indicator calcultions using `demogsurv`}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette describes the methods implemented by `demogsurv` for calculating demographic indicators.

## Setup

Load the `demogsurv` package.

```{r}
## devtools::install_github("mrc-ide/demogsurv")
library(demogsurv)
library(survival)
library(survey)
```

[DHS model datasets](https://dhsprogram.com/data/download-model-datasets.cfm) are included as example datsets in the package (see `?zzir`). Load the example datasets:

```{r load-datasets}
data(zzir) # Individual recode
data(zzbr) # Birth's recode
```

# Child mortality (~5~q~0~)

The package implemets a generic function `calc_nqx()` for calculating the 
probability of dying between ages $x$ and $x+n$. For calculation of ~5~q~0~, 
`calc_nqx()` proceeds by:

1. Calculating the numbers deaths and person-years observed within the age ranges
   0, 1-2, 3-5, 6-11, 12-23, 24-35, 36-47, and 48-60 months.
1. Calculating the mortality rate within each age group.
1. Aggregate mortality rates to cumulative mortality hazard over age 0 to 4 years.
1. Convert cumulative hazard to probability of dying.

The calculation uses the same age group stratification as the standard DHS indicators
reported for infant and child mortality. But the calculation differs in that it 
calculates a piecewise constant mortaity hazard within each of these age groups
and converts the cumulative hazard to probability of dying, whereas the DHS indicator
calculates the probability of dying within each age group as the ratio of the number 
of deaths divided by the number surviving to the start of the age group.
calculates the product overall ages (see [Rutstein and Rojas 2006](http://dhsprogram.com/pubs/pdf/DHSG1/Guide_to_DHS_Statistics_29Oct2012_DHSG1.pdf)).

## Define survival episodes for each child born

The calculation uses the _Birth's recode_ dataset (`zzbr`), which is formatted as 
one row per live born child reported in the full birth history. The first step is 
to define survival episodes for each child.

Define a binary variable indicating whether a death occurred or the episode is 
censored.
```{r event_var}
zzbr$death <- zzbr$b5 == "no"  # b5: child still alive (1 = "yes", 2 = "no")
```

Variable `b7` reports the child age at death in completed months. The date of
death is calculated as the date of birth plus the age at death plus 0.5 months,
such that on average one half month of person-time is contributed in the month
of death.
```{r dod}
zzbr$dod <- zzbr$b3 + zzbr$b7 + 0.5
```

Survivor episodes start at the child date of birth (variable `b3`) and ends at 
either the child date of death or are right  censored at the date of interview 
(`v008`).
```{r censor}
zzbr$tstop <- ifelse(zzbr$death, zzbr$dod, zzbr$v008)
```

## Aggregate weighted numbers of deaths and person years by age group

The function `demog_pyears()` is a wrapper for the function `survival::pyears()` to 
calculate numbers of events (deaths) and person years stratfied by age group (`agegr`), 
calendar period (`period`), time preceding the survey (`tips`), or birth cohort (`cohort`)
from survival episode data. We use `demog_pyears()` to calculate the number of child
deaths and person years in the age groups 0, 1-2, 3-5, 6-11, 12-23, 24-35, 36-47, 
and 48-59 months during the period 0-4 years before the survey.

```{r, demog_pyears, }
zzbr$weight <- zzbr$v005 / mean(zzbr$v005) # Normalize weights
aggr <- demog_pyears(~1, zzbr, agegr=c(0, 1, 3, 6, 12, 24, 36, 48, 60)/12, tips=c(0, 5),
                     event="death", tstart="b3", tstop="tstop",
                     dob = "b3", intv = "v008", weights="weight",
                     origin = 1900, scale = 12)
```

## Calculate mortality rate within each age group

The mortality rate $mu_a$ within each age group is estimated as the number
of observed deaths divided by the number of person-years.

```{r mx}
mx <- aggr$data
mx$agegr <- paste(c(0, "1-2", "3-5", "6-11", "12-23", "24-35", "36-47", "48-59"), "months")
mx$mx <- mx$event / mx$pyears
knitr::kable(mx)
```

## Cumulative hazard and probability of dying

The cumualative hazard is the sum over all age groups of the mortality rate
in each age group times the span of the age group.
```{r cumhaz}
mx$a_min <- c(0, 1, 3, 6, 12, 24, 36, 48) / 12
mx$a_max <- c(1, 3, 6, 12, 24, 36, 48, 60) / 12

cum_mx <- sum( (mx$a_max - mx$a_min) * mx$mx)
cum_mx
```

Finally, the probability of dying ~5~q~0~ is one minus the cumulative probability
of escaping death:
```{r 5q0}
1 - exp(-cum_mx) # 5q0
```

## Standard error estimation 

### Via Taylor linearization 

_To be completed_

### Via stratified jackknife 

_To be completed_


# Adult mortality (~35~q~15~, ~45~q~15~)

_To be completed_

# Total fertility rate

_To be completed_

