---
title: "Process and calculate demographic rates from DHS datasets"
author: "Jeff Eaton, Bruno Masquelier"
date: "`r Sys.Date()`"
output:
  html_document: 
    smart: false
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette illustrates use of the `demogsurv` package to:
* process Demographic and Health Survey (DHS) datasets,
* replicate the standard demographic indicators produced in DHS reports, and
* calculate indicators for customized periods or age groups.

## Load DHS survey datasets

DHS provides an example model dataset downloadable without registration from [here](https://dhsprogram.com/data/download-model-datasets.cfm). Download the Stata dataset version of the individual recode and birth recode datasets into temporary files.

```{r}
## irdt_zip <- tempfile()
## download.file("https://dhsprogram.com/customcf/legacy/data/sample_download_dataset.cfm?Filename=ZZIR62DT.ZIP&Tp=1&Ctry_Code=zz&survey_id=0&doctype=dhs", irdt_zip)
## brdt_zip <- tempfile()
## download.file("https://dhsprogram.com/customcf/legacy/data/sample_download_dataset.cfm?Filename=ZZBR62DT.ZIP&Tp=1&Ctry_Code=zz&survey_id=0&doctype=dhs", brdt_zip)
```

The function `read_zipdta()` identifies and reads the file with extension `.DTA` from within the zip archive. `read_zipdta()` is a wrapper for the more generic function `read_zipdata()` which can match an arbitrary filename inside a zip directory using regular expressions (argument `pattern`) and read the file using a given function (argument `readfn`).

```{r, warning=FALSE}
## ir <- read_zipdta(irdt_zip)
## br <- read_zipdta(brdt_zip)
```

## Adult and maternal mortality

WORK IN PROGRESS
