---
title: "Using `demogsurv` with `rdhs`"
author: "Jeff Eaton, Bruno Masquelier, and OJ Watson"
date: "`r Sys.Date()`"
output:
  html_document: 
    smart: false
    keep_md: true
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

This vignette illustrates use of `demogsurv` and `rdhs` to calculate fertility and mortality indicators for lots of DHS surveys in sub-Saharan Africa, and compare estimates to those produced for the [DHS StatCompiler](https://www.statcompiler.com/en/). It is currently a hastily developed analysis script, though may be further developed in the future.

## Install and load packages

```{r install}
## install.packages("devtools")
## devtools::load_all("OJWatson/rdhs")
## devtools::load_all("mrc-ide/demogsurv")

library(rdhs)
library(demogsurv)
library(ggplot2)
library(data.table)
library(haven)

## a little nugget to return API requests as data.table rather than data.frame.
Sys.setenv(rdhs_DATA_TABLE = "TRUE")
```

## Identify surveys and datasets

Identify all DHS surveys conducted in sub-Saharan Africa since the year 2005.

```{r dhs_surveys}
countries <- dhs_countries()
cc <- countries[RegionName == "Sub-Saharan Africa"]$DHS_CountryCode
surveys <- dhs_surveys(countryIds = cc, surveyYearStart=2005, surveyType = "DHS")
```

Identify individual recode (IR) and births recode (BR) datasets corresponding to these surveys.

```{r dhs_datasets}
ird <- dhs_datasets(fileType = "IR", fileFormat = "flat")[SurveyId %in% surveys$SurveyId]
brd <- dhs_datasets(fileType = "BR", fileFormat = "flat")[SurveyId %in% surveys$SurveyId]
```

Use `rdhs` to retreive datasets, downloading them from DHS website if not already in the `rdhs` cache.

```{r get_datasets, warning=FALSE, message=FALSE}
ird$path <- unlist(get_datasets(ird$FileName))
brd$path <- unlist(get_datasets(brd$FileName))
```

Load all of the datasets into R as a list.

```{r load datasets, results="hide", eval=FALSE}
ir <- list()
for(survid in ird$SurveyId){
  print(survid)
  dat <- readRDS(ird[SurveyId == survid]$path)
  dat <- dat[grep("caseid|^v0|^v1|^b|^mm", names(dat))]
  ir[[survid]] <- dat
}

br <- list()
for(survid in brd$SurveyId){
  print(survid)
  dat <- readRDS(brd[SurveyId == survid]$path)
  dat <- dat[grep("caseid|^v0|^v1|^b", names(dat))]
  br[[survid]] <- dat
}

## Convert to factors (a bit inefficient)
ir <- lapply(ir, haven::as_factor)
br <- lapply(br, haven::as_factor)

## Add survey-level variables
ir <- Map(data.frame,
          SurveyId = surveys$SurveyId,
          CountryName = surveys$CountryName,
          SurveyYear = surveys$SurveyYear,
          ir)

br <- Map(data.frame,
          SurveyId = surveys$SurveyId,
          CountryName = surveys$CountryName,
          SurveyYear = surveys$SurveyYear,
          br)
```

Note that `rdhs` provides better tools to extract variables and pool datasets which haven't been fully embraced here.


## Use `demogsurv` to analyse demographic rate indicators

### Fertility

Calcualte TFR and 15-19 ASFR for 3 year period preceding survey (default argument `tips=c(0, 3)`).
```{r tfr, eval=FALSE}
tfr <- lapply(ir, calc_tfr, by=~SurveyId+CountryName+SurveyYear, strata=NULL)
tfr <- do.call(rbind, tfr)

asfr15to19 <- lapply(ir, calc_asfr, by=~SurveyId + CountryName + SurveyYear,
                     agegr = c(15, 20), strata=NULL)
asfr15to19 <- do.call(rbind, asfr15to19)
```


### Adult mortality

Identify surveys that include sibling history model via querying the DHS API survery with "Maternal mortality" characteristic.

```{r has sibling}
survchar <- dhs_surveyCharacteristics()
survchar[grepl("Maternal", SurveyCharacteristicName)]
mm_surv <- dhs_surveys(surveyCharacteristicIds = 1)
has_mm <- surveys$SurveyId %in% mm_surv$SurveyId
```

Reshape IR datasets to one row per sibling episode, create a binary variable indicating sibling death, 
and calculate ~35~q~15~ estimates by sexx.

```{r sibmx, warning=FALSE, eval=FALSE}
sib <- lapply(ir[has_mm], reshape_sib_data,
              widevars = c("SurveyId", "CountryName", "SurveyYear", "v005", "v008", "v021"))
sib <- lapply(sib, function(x){x$death <- factor(x$mm2, c("dead", "alive")) == "dead"; x})
q3515 <- lapply(sib, calc_nqx, by=~SurveyId+CountryName+SurveyYear + mm1, strata = NULL,
                agegr=seq(15, 50, 5), tips=c(0, 7), dob="mm4", dod="mm8")
q3515 <- do.call(rbind, q3515)
```	

### Child mortality

`demogsurv` does not yet implement the exact child mortality calculation produced in DHS
reports and DHS StatCompiler (see [Rutstein and Rojas 2006](http://dhsprogram.com/pubs/pdf/DHSG1/Guide_to_DHS_Statistics_29Oct2012_DHSG1.pdf).
This is planned for future implementation. 


The function `calc_nqx()` calculates piecewise constant mortality rates within 
age groups 0, 1-2, 3-4, 5-11, 12-24 months, and 2, 3, and 4-5 years (parameter
`agegr = c(0, 1, 3, 5, 12, 24, 36, 48, 60)/12`).  These are aggregated to a cumulative 
hazards over the age group 0-4 years and converted to probabilities to estimate ~5~q~0~.

Add a binary indicator whether a death occurred and a date of death variable, placed 0.5 
months in the month the death occurred.

```{r deathvar, eval=FALSE}
br <- lapply(br, function(x){x$death <- x$b5 == "no"; x})
br <- lapply(br, function(x){x$dod <- x$b3 + x$b7 + 0.5; x})
```

Calculate ~5~q~0~ for period 0-4, 5-9, and 10-14 years preceding the survey.

```{r 5q0, eval=FALSE}
u5mr <- lapply(br, calc_nqx, by=~SurveyId+CountryName+SurveyYear, strata=NULL)
u5mr <- do.call(rbind, u5mr)
```


## Merge DHS StatCompiler indicators

```{r sneaky load, echo=FALSE, results="hide"}
tfr <- readRDS("rdhs-integration/tfr.rds")
asfr15to19 <- readRDS("rdhs-integration/asfr15to19.rds")
q3515 <- readRDS("rdhs-integration/q3515.rds")
u5mr <- readRDS("rdhs-integration/u5mr.rds")
```

Identify the indicator IDs associated with TFR, ASFR 15-19, ~35~q~15~, and ~5~q~0~.

```{r dhs_indicators}
indic <- dhs_indicators()

indic[grepl("TFR 15-49", ShortName), .(IndicatorId, ShortName, Label)]
indic[grepl("ASFR 15-19", ShortName), .(IndicatorId, ShortName, Label)]
indic[grepl("Probability of dying", ShortName), .(IndicatorId, Definition)]
indic[grepl("Under-five mortality", ShortName), .(IndicatorId, Label)]
```

Query estimates from DHS API and merge with calculated estimates.

```{r dhs_data}

tfr_dhs <- dhs_data(indicatorIds = "FE_FRTR_W_TFR",
                    surveyId = tfr$SurveyId)
tfr <- merge(tfr, tfr_dhs[ , .(SurveyId, Value)])

asfr15to19_dhs <- dhs_data(indicatorIds = "FE_FRTR_W_A15",
                           surveyId = asfr15to19$SurveyId)
asfr15to19 <- merge(asfr15to19, asfr15to19_dhs[ , .(SurveyId, Value)])

q3515_dhs <- dhs_data(indicatorIds = c("MM_AMPB_W_AMP", "MM_AMPB_M_AMP"),
                      surveyId = q3515$SurveyId)
q3515_dhs$mm1 <- c(MM_AMPB_M_AMP = "male", MM_AMPB_W_AMP = "female")[q3515_dhs$IndicatorId]
q3515 <- merge(q3515, q3515_dhs[ , .(SurveyId, mm1, Value)])

u5mr_dhs <- dhs_data(indicatorIds = "CM_ECMT_C_U5M",
                     surveyYearStart = 2005,
                     breakdown = "all")[SurveyId %in% u5mr$SurveyId]
u5mr_dhs$tips <- u5mr_dhs$CharacteristicLabel
u5mr <- merge(u5mr, u5mr_dhs[, .(SurveyId, tips, Value)])
```

## View estimates

```{r review estimates, echo=FALSE}
knitr::kable(head(tfr), digits=c(rep(0, 4), 1, 3, 2),
             caption = "TFR")

knitr::kable(head(asfr15to19), digits=c(rep(0, 5), 3, 4, 0),
             caption = "ASFR 15-19")

knitr::kable(head(q3515), digits=c(rep(0, 5), rep(3, 4), 0),
             caption = "35q15")

hu5mr <- head(u5mr)
hu5mr$tips <- factor(hu5mr$tips, c("0-4", "5-9", "10-14"))
hu5mr <- hu5mr[order(hu5mr$SurveyId, hu5mr$tips),]
knitr::kable(hu5mr, digits=c(rep(0, 4), rep(3, 4), 0),
             caption = "5q0")
```

## Check that TFR, ASFR, and ~35~q~15~ estimates exactly match

Estimates for fertility rates and adult mortality rates should exactly match
those produced as standard DHS indicators.


```{r check match}

## TFR matches exactly
with(tfr, table(round(tfr, 1) == Value))

## ASFR 15-19 matches exactly
with(asfr15to19, table(round(1000*asfr) == Value))

## 35q15 matches exactly for >80%
with(q3515, table(round(1000*est) == Value))
with(q3515, table(round(1000*est) - Value))

subset(q3515, abs((round(1000*est) - Value)) > 1)
```


## Compare ~5~q~0~ estimates

```{r 5q0 compare, fig.height=3, fig.width=4.5, fig.align="center"}
u5mr$tips <- factor(u5mr$tips, c("0-4", "5-9", "10-14"))
ggplot(u5mr, aes(1000*est, Value, color=tips)) +
  geom_abline(slope=1, color="grey") +
  geom_point() +
  coord_fixed() +
  xlab("demogsurv::calc_nqx()") +
  ylab("DHS StatCompiler") +
  ggtitle("5q0 comparison")
```
